---
title: "Continuous glucose monitoring for metabolic profiling in prediabetes"
subtitle: "DDEA application"
author: "Mikkel Errboe Steentoft"
date: 2026-08-05
date-format: "DD-MM-YYYY"
editor: visual
bibliography: references.bib
link-citations: true
csl: https://zotero.org/styles/diabetologia

format:
  docx:
    toc: false
    number-sections: true
  typst:
    toc: false
    number-sections: true
    page-numbering: "1"
---

## Background

Prediabetes is a state of elevated blood glucose levels below the threshold for type 2 diabetes and is associated with increased risk of diabetes [@tabák2012], cardiovascular disease (CVD) and mortality [@vistisen2018]. Prediabetes is currently defined by different glycemic thresholds; elevated glycated hemoglobin (HbA1c) level of 39-47 mmol/mol [@americandiabetesassociationprofessionalpracticecommittee2025] or 42-47 mmol/mol [@internat2009], elevated fasting glucose levels (IFG) or elevated 2-hour-glucose (2hPG) after an oral glucose tolerance test (OGTT). While current diagnostic criteria for prediabetes capture the degree of dysglycemia at a single time point, they offer limited insight into underlying pathophysiology. Individuals meeting the same glycemic criteria for prediabetes may differ substantially in their underlying metabolic drivers and risk trajectories. This includes distinct characteristics of insulin resistance, beta-cell function and body fat distribution along with varying risks of developing diabetes and related complications [@wagner2026]. Given this substantial heterogeneity among individuals with prediabetes, effective prevention requires the ability to distinguish high-risk individuals who could benefit from early, intensive intervention from low-risk individuals who require minimal clinical attention [@blond2023]. However, assessment of the underlying metabolic drivers currently relies on complex tools, such as the OGTT for measuring post-load glucose regulation, or MRI and ultrasonography for the assessment of body fat distribution [@wagner2021; @stroebel2025]. An important step towards precision prevention is therefore to develop more clinically accessible tools capable of capturing underlying metabolic dimensions.

Continuous glucose monitoring (CGM) offers a possibility to assess detailed and dynamic indices of glucose regulation under free-living conditions, and is already widely used in the management of type 1 and type 2 diabetes [@battelino2025]. Emerging evidence suggests that the use of CGM could also be beneficial in the management of individuals with prediabetes [@zahalka2024]. Recent studies including large-scale machine learning models have shown that CGM- and wearable-derived measures can predict future glycemic deterioration, incident type 2 diabetes, cardiovascular outcomes [@lutsker2026] and insulin resistance [@metwally2026]. While these approaches have the potential to provide important information for risk stratification, they provide limited insight into the distinction between insulin resistance and beta-cell function - a distinction that may be important not only for identifying who is at risk, but also for determining which type of preventative strategy, such as intensive lifestyle interventions or early pharmacological treatment, is most suitable for a given individual [@wagner2026]. In this context, it remains unclear whether CGM- and wearable-derived information in free-living conditions can help identify underlying metabolic dimensions, such as insulin resistance and beta-cell function, among individuals with prediabetes, and how this could inform future precision prevention strategies.

## Aims

The overall aim of this PhD project is to assess whether continuous glucose monitoring, combined with other free-living data, can help characterize underlying metabolic dimensions, including OGTT-derived measures of insulin resistance and beta-cell function, among individuals with prediabetes. The project comprises three studies, encompassing cross-sectional and longitudinal observational analyses and detailed physiological characterisation. Study 1 establishes whether CGM- and lifestyle-derived measures are associated with insulin resistance and beta-cell function in a prediabetes cohort. Study 2 examines whether these associations are observed in an independent cohort and relate to long-term outcomes. Study 3 provides physiological insight into the metabolic processes reflected by CGM response patterns. Table 1 provides an overview for study aims, data sources and populations across the three studies.

```{r}
#| label: tbl-studies
#| tbl-cap: "Overview of study aims, data sources and populations"
#| echo: false
#| message: false
library(flextable)
library(tibble)

studies <- tribble(
  ~Study, ~Aim, ~`Data source`, ~ Population,

  "Study 1 Cross-sectional",
  "Assess how CGM-derived glycemic patterns, along with free-living lifestyle data (physical activity, sleep, diet), are associated with insulin resistance and beta-cell function among individuals with prediabetes.",
  "Baseline data from the DP-NEXT study (Work Package 3: Heterogeneity), a Danish study characterizing heterogeneity among individuals with prediabetes.",
  "Danish adults aged 40-50 years with an HbA1c measurement between 42-47 mmol/mol.",
  
  "Study 2 Longitudinal",
  "Investigate how CGM-derived glycemic patterns and related free-living lifestyle characteristics are associated with insulin resistance and beta-cell function, along with long-term outcomes including type 2 diabetes, cardiovascular disease and mortality.",
  "Baseline and follow-up data from the Maastricht Study, a Dutch phenotyping study focusing on the etiology of type 2 diabetes.",
  "Dutch adults aged 40-75 years with an HbA1c measurement between 42-47 mmol/mol.",
  
  "Study 3 Exploratory, hypothesis-generating physiological study",
  "Characterize underlying metabolic dimensions, including insulin resistance, beta-cell function and substrate metabolism, and how they relate to CGM response patterns.",
  "Exploratory physiological study using whole-room indirect calorimetry at Steno Diabetes Center Aarhus.",
  "5-10 participants with prediabetes from Work Package 3 in the DP-NEXT study."
)

ft <- flextable(studies) |>
  theme_vanilla() |>
  bold(j = "Study", bold = TRUE) |>
  width(j = "Study", width = 0.9) |>
  width(j = "Aim", width = 2.3) |>
  width(j = "Data source", width = 2.0) |>
  width(j = "Population", width = 1.8) |>
  valign(valign = "top", part = "body") |>
  fontsize(size = 10, part = "all") |>
  font(fontname = "Calibri", part = "all") |>
  padding(padding = 4, part = "all")

ft
```

## Research design and methods

**Study 1:** Cross-sectional study of the associations between CGM-derived glucose patterns, free-living measures of physical activity, sleep, diet and oral glucose tolerance test-derived indices of insulin resistance and beta-cell function.

The aim of work package 3 in the DP-NEXT study is to characterize heterogeneity among individuals with prediabetes and will collect a large variety of data (<https://dp-next.github.io/wp3.html>). This study will assess which combinations of free-living clinically accessible measures, separate or in combinations, best capture variation in insulin resistance and beta-cell function derived from the oral glucose tolerance test. In this study, we will use the following data from the DP-NEXT study (but not limited to): Free-living dynamic glucose patterns through CGM, physical activity and sleep through wrist-worn actigraphy, diet registration, glucose and insulin measures from a 4-point OGTT, sociodemographic characteristics and anthropometric characteristics and measures of body fat distribution.

**Study 2:** Longitudinal study assessing associations between baseline CGM-derived glucose patterns, related lifestyle characteristics, insulin resistance and beta-cell function, as well as their relevance in predicting long-term outcomes including type 2 diabetes, cardiovascular disease and mortality.

The Maastricht Study is a large phenotyping cohort study investigating the etiology of type 2 diabetes, with baseline and follow-up data available for a subset of participants [@schram2014]. In this study, we will use the following data from the Maastricht Study at baseline and follow-up where available (but not limited to): Free-living dynamic glucose patterns through CGM, physical activity and sleep through accelerometry, diet registration, glucose and insulin measures from an OGTT, sociodemographic characteristics, anthropometric characteristics, measures of body fat distribution and incident type 2 diabetes, cardiovascular disease and mortality.

**Study 3:** Exploratory physiological study characterizing underlying metabolic dimensions of prediabetes including insulin resistance, beta-cell function and substrate metabolism and their relation to CGM patterns under controlled conditions.

A small number (5-10) of participants from the DP-NEXT study (see study 1), selected to represent high-risk metabolic profiles will be invited to undergo detailed physiological characterization using the whole-room indirect calorimetry room at Steno Diabetes Center Aarhus. In this study, we will collect the following data (but not limited to): Indirect calorimetry measures of energy metabolism at rest and in response to diet and exercise protocols, glucose and insulin measures from an OGTT and dynamic glucose patterns through CGM.

## References

```{r}
#| echo: false
#| message: false
#refs
```
